---
title: "Trade Presentation"
author: "Shahzia Teja"
date: "July 8, 2019"
output: 
  revealjs::revealjs_presentation:
    theme: blood
    css: style.css
    self_contained: false
    reveal_plugins: ["chalkboard"]
reveal_options:
  slideNumber: true
  previewLinks: true
  chalkboard:
    theme: whiteboard
    toggleNotesButton: false
  slide_level: 2
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
options(stringsAsFactors = FALSE, scipen = 20)
library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)

```

# Ceteris Paribus {data-background-iframe="https://vi.unctad.org/tpa/web/vol1/vol1home.html"}

## Partial Equilibrium

Partial Equilibrium for Trade & Regional Analysis

Partial Equilibrium models allow us to examine the effects of economic shocks in a parsimonious modeling environment with few data requirements or assumptions.

In this world, a specific market clears without regard to prices and quantity changes in other markets. In other words, all other prices, quantities and income are assumed to be fixed. Some of the advantages of this approach are obvious, including the ability to create and run models quickly, as well as producing intuitively appealing results.

## Tariffs as a Shock

When countries impose trade barriers, in this case in the form of tarrifs, prices and quantities of implicated goods can be expected to respond.

We assume a large open economy with downward sloping import demand curves (this implies the country's demand affects prices). In general, an importing country is assumed to have prices above the world price, since the alternative would imply no trade takes place. Buyers are assumed to be indifferent between domestic and foreign-produced substitutes.

# Simplest Partial Equilibrium Model

## System Functions

Domestic Demand: $Q_d=\alpha(P)^\eta$

Domestic Supply: $Q_s=\beta(P)^\epsilon$

Import Demand (the excess of domestic demand over domestic supply): $M_d=\alpha(P)^\eta-\beta(P)^\epsilon$

Import Supply: $M_s=Y(P_w)^\theta$

Internal Price: $P=(1+t)*P_w$

## Autarky

Without imports, domestic demand equals domestic supply.

$Q_d = Q_s$ or $\alpha P^\eta = \beta P^\epsilon$

which implies $\alpha / \beta = P^{\epsilon - \eta}$

This can be solved for:
$$P_0 = (\alpha / \beta) ^ \frac{1}{\epsilon - \eta}$$
$$Q_0 = \alpha P_0^\eta = \alpha (\alpha / \beta) ^ \frac{\eta}{\epsilon - \eta} = \alpha ^ \frac{\epsilon}{\epsilon - \eta} \beta ^ \frac{-\eta}{\epsilon - \eta}$$

## Free Trade

In this case, import demand equals import supply...

$M_d = M_s$ or $\alpha P^\eta - \beta P^\epsilon = \gamma P_w^\theta$

but $P = (1 + t) P_w$ which implies:

$\alpha (1 + t)^\eta P_w^\eta - \beta (1 + t)^\epsilon P_w^\epsilon = \gamma P_w^\theta$


## Autarky Illustrated

```{r echo=FALSE, warning=FALSE, message=FALSE}
# demand and supply
Qd <- function(p=1, alpha = 100, eta = -1) {
  alpha * p ^ eta
}

Pd <- function(q = 100, alpha = 100, eta = -1) {
  (q/alpha)^(1/eta)
}

Qs <- function(p = 1, beta = 100, epsilon = 3) {
  beta * p ^ epsilon
}


par_alpha <- 7480.06
par_beta  <- 1182.38
par_gamma <- par_alpha - par_beta

par_eta     <- -1
par_epsilon <- 3
par_theta   <- 10

# no trade eq'm
p0 <- (par_alpha / par_beta) ^ (1 / (par_epsilon - par_eta))
q0 <- par_alpha * p0 ^ par_eta

# benchmark equilibrium
pars <- c(a = par_gamma, b = par_theta - par_eta, c= par_beta, d = par_epsilon - par_eta, e = par_alpha)

f_bm <- function(pw, pars) {
  unname(pars['a'] * pw ^ pars['b'] + pars['c'] * pw ^ pars['d'] - pars['e'])
}

rslt <- uniroot(f_bm, c(0.1,2), pars, tol=5e-10)
pw <- rslt$root

qw <- par_alpha * pw ^ par_eta
qw_d <- par_beta * pw ^ par_epsilon
qw_i <- qw - qw_d
```

Suppose: 

$\alpha =$ `r par_alpha`, $\beta =$ `r par_beta`, $\gamma =$ `r par_gamma`,
$\eta =$ `r par_eta`, $\epsilon =$ `r par_epsilon` and $\theta =$ `r par_theta` 

Ignoring tariffis for now...

This implies $P_0 =$ `r round(p0,2)` and $Q_0 =$ `r round(q0,2)`

Whereas in the benchmark case, 

$P^* = P_w = 1$ and $Q^* = Q_{Domestic} + Q_{Imported}$

= `r par_beta` + `r par_gamma` = `r par_alpha`

----

```{r echo=FALSE, warning=FALSE, message=FALSE}
p <- seq(0.01, 2, .01)
qd <- Qd(p, par_alpha)  # demand
qs <- Qs(p, par_beta)  # domestic supply
md <- qd - qs         # import demand

mx <- max(c(qd, qs))
mn <- min(c(qd, qs))

df1 <- data.frame(x = qd, y = p, curve = rep("demand", length(p)))
df2 <- data.frame(x = qs, y = p, curve = rep("supply", length(p)))
df3 <- data.frame(x = md, y = p, curve = rep("imports", length(p)))

df <- rbind(df1, df2, df3[df3$x >= 0,])

plt1 <- ggplot(df, aes(x=x,y=y,color=curve)) + 
  geom_line(size=1.5) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  annotate("segment", x = 0, xend = q0, y = p0, yend = p0, linetype = 3) +
  annotate("segment", x = q0, xend = q0, y = p0, yend = 0, linetype = 3) +
  annotate("segment", x = 0, xend = par_alpha, y = 1, yend = 1, linetype = 2) +
  annotate("segment", x = par_beta, xend = par_beta, y = 1, yend = 0, linetype = 2) +
  annotate("segment", x = qw_i, xend = qw_i, y = 1, yend = 0, linetype = 2) +
  annotate("segment", x = par_alpha, xend = par_alpha, y = 1, yend = 0, linetype = 2) +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0,2)) +
  xlab("quantity") +
  ylab("price") +
  labs(title="Demand, Supply and Import Demand",
       subtitle = "Import Demand is residual demand") +
  annotate("text", x=q0, y=0.1, label="q[0]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=1.05, label="p[w]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=p0 + .05, label="p[0]", parse = TRUE, size = 5.2) +
  annotate("text", x = par_beta, y=0.1, label = "q[Dom]", parse = TRUE, size = 5.2) +
  annotate("text", x = qw_i, y=0.1, label = "q[Imp]", parse = TRUE, size = 5.2) +
  annotate("text", x = par_alpha, y=0.1, label = "q[Tot]", parse = TRUE, size = 5.2)
print(plt1)  

```

## Impact of Moving from Autarky to Free Trade

Using demand and supply curves defined above, price falls from $P_0$ = `r round(p0, 2)` to $P_w$ = `r round(pw,2)`

## Consumer Surplus

$\Delta CS = -\int_{P_0}^{P_w} Q_d(u) du$

but $p_w$ < $p_0$, so, 

$\Delta CS = -\int_{P_w}^{P_0} -Q_d(u) du = \alpha \int_{P_w}^{P_0} u^\eta du$

if $\eta \neq -1$, this equals $\alpha \left[\frac{u^{\eta + 1}}{\eta + 1}\right]_{P_w}^{P_0}$ 

if $\eta = -1$, this equals $\alpha \left[\ln u \right]_{P_w}^{P_0}$


In this case, $\Delta CS =$ `r round(par_alpha * (log(p0) - log(pw)),2)`

----

```{r echo=FALSE, warning=FALSE, message=FALSE}
df <- rbind(df1, df2)

df4 <- data.frame(x = qd, y = p, curve="demand")
df4 <- subset(df4, x <= qw & (y >= pw & y <= p0))
df4 <- rbind(df4, df4[1:2,])
row.names(df4) <- 1:nrow(df4)
df4[nrow(df4)-1, 1] <- 0
df4[nrow(df4), 1] <- 0
df4[nrow(df4)-1, 2] <- p0
df4[nrow(df4), 2] <- 1

plt2 <- ggplot(df, aes(x=x,y=y,color=curve)) + 
  geom_line(size=1.5) +
  geom_polygon(data = df4, aes(x=x,y=y), color="grey", alpha=0.2, show.legend = FALSE) +
  geom_hline(yintercept = c(0, pw, p0)) +
  geom_vline(xintercept = 0) +
  annotate("segment", x= 350, xend = 350, y = p0 - 0.01, yend = pw + 0.01, arrow=arrow(length = unit(0.2, "cm"))) +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0,2)) +
  annotate("text", x=350, y=p0 + .05, label="p[0]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=pw - .05, label="p[w]", parse = TRUE, size = 5.2) +
  xlab("quantity") +
  ylab("price") +
  labs(title = "Increase in Consumer Surplus",
       subtitle = "Opening to trade makes consumers better off")

print(plt2)
```

## Producer Surplus

$\Delta PS = \int_{P_0}^{P_w} Q_s(u) du$

but $p_w$ < $p_0$, so, 

$\Delta PS = -\int_{P_w}^{P_0} Q_s(u) du = -\beta \int_{P_w}^{P_0} u^\epsilon du$

Since $\epsilon > 0$, this equals $-\beta \left[\frac{u^{\epsilon + 1}}{\epsilon + 1} \right]_{P_w}^{P_0}$


In this case, $\Delta PS =$ `r round(-par_beta * (p0^(par_epsilon+1) - pw^(par_epsilon+1))/(par_epsilon+1),2)`

----

```{r echo=FALSE, warning=FALSE, message=FALSE}
# producer surplus
df5 <- data.frame(x = qs, y = p, curve="supply")
df5 <- subset(df5, x <= qw & (y >= pw & y <= p0))
df5 <- rbind(df5, df5[1:2,])
row.names(df5) <- 1:nrow(df5)
df5[nrow(df5)-1, 1] <- 0
df5[nrow(df5), 1] <- 0
df5[nrow(df5)-1, 2] <- p0
df5[nrow(df5), 2] <- 1

plt3 <- ggplot(df, aes(x=x,y=y,color=curve)) + 
  geom_line(size=1.5) +
  geom_polygon(data = df5, aes(x=x,y=y), color="grey", alpha=0.2, show.legend = FALSE) +
  geom_hline(yintercept = c(0, pw, p0)) +
  geom_vline(xintercept = 0) +
  annotate("segment", x= 350, xend = 350, y = p0 - 0.01, yend = pw + 0.01, arrow=arrow(length = unit(0.2, "cm"))) +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0,2)) +
  annotate("text", x=350, y=p0 + .05, label="p[0]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=pw - .05, label="p[w]", parse = TRUE, size = 5.2) +
  xlab("quantity") +
  ylab("price") +
  labs(title = "Reduction in Producer Surplus",
       subtitle = "Opening to trade makes producers worse off")

print(plt3)
```

## Net Change in Welfare

In this case we have $\Delta CS + \Delta PS > 0$

In fact, $\Delta CS + \Delta PS =$ `r round( par_alpha * (log(p0) - log(pw))- par_beta * (p0^(par_epsilon+1) - pw^(par_epsilon+1))/(par_epsilon+1), 2)`

----

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Net gain in welfare
x <- seq(pw, p0, .01)
y <- Qd(x, par_alpha)
x1 <- seq(p0 - 0.01, pw, -0.01)
y1 <- Qs(x1, par_beta)
df6 <- data.frame(x = c(y,y1), y = c(x,x1), curve="demand")

plt4 <- ggplot(df, aes(x=x,y=y,color=curve)) + 
  geom_line(size=1.5) +
  geom_polygon(data = df6, aes(x=x,y=y), color="grey", alpha=0.2, show.legend = FALSE) +
  geom_hline(yintercept = c(0, pw, p0)) +
  geom_vline(xintercept = 0) +
  annotate("segment", x= 350, xend = 350, y = p0 - 0.01, yend = pw + 0.01, arrow=arrow(length = unit(0.2, "cm"))) +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0,2)) +
  annotate("text", x=350, y=p0 + .05, label="p[0]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=pw - .05, label="p[w]", parse = TRUE, size = 5.2) +
  xlab("quantity") +
  ylab("price") +
  labs(title = "Net Increase in Welfare",
       subtitle = "Consumers more better off than producers are worse off")

print(plt4)
```

# Impact of Tariff in Simple Model

## Prices \& Quantities

```{r echo=FALSE, warning=FALSE, message=FALSE}
# tariff equilibrium
t <- 0.2
pars <- c(a = par_gamma, 
          b = par_theta - par_eta, 
          c = par_beta * (1 + t) ^ par_epsilon, 
          d = par_epsilon - par_eta, 
          e = par_alpha * (1 + t) ^ par_eta)

rslt <- uniroot(f_bm, c(0.1,2), pars, tol=5e-10)
pw1 <- rslt$root

p1 <- (1 + t) * pw1
q1 <- par_alpha * p1 ^ par_eta
q1_d <- par_beta * p1 ^ par_epsilon
q1_i <- q1 - q1_d

v_before <- c(pw, pw, qw_i, qw_d, qw, 0)
v_after  <- c(pw1, p1, q1_i, q1_d, q1, (p1 - pw1) * q1_i)
v_change <- v_after - v_before
v_pct    <- ifelse(v_before > 0, v_change / v_before * 100, NA)
v_decs   <- c(3, 3, 2, 2, 2, 2)

c1 <- mapply(function(x,y) format(round(x,y), nsmall = y), v_before, v_decs)
c2 <- mapply(function(x,y) format(round(x,y), nsmall = y), v_after, v_decs)
c3 <- mapply(function(x,y) format(round(x,y), nsmall = y), v_change, v_decs)
c4 <- format(round(v_pct,2),nsmall = 2)

df_tab <- data.frame(Variables = c("Pw", "P", "Md", "Qs", "Qd", "Tariffs"),
                     Before = c1, After =  c2, Change = c3, Percent = c4)
kable(df_tab)

TTE <- -0.5 * (qw_i + q1_i) * (pw1 - pw) # Terms of Trade Effect

HTE <- 0.5 * (p1 - pw1) * (q1_i - qw_i) # Harberger Triangle Effect

WE  <- TTE + HTE # Welfare Effect

dCS <- -par_alpha * (log(p1) - log(pw))
dPS <- par_beta * (p1^(par_epsilon + 1) - pw^(par_epsilon + 1)) / (par_epsilon + 1)
dTariffs <- (p1 - pw1) * q1_i
```

$dW = -M dP_w + \frac{1}{2}(P - P_w) dM$

Terms of trade effect: $-M dP_w =$ `r round(TTE,2)`

Harberger triangle effect: $\frac{1}{2}(P - P_w) dM =$ `r round(HTE,2)`

Welfare effect: $dW =$ `r round(TTE + HTE, 2)`

## How Accurate is the approximation?

$dW =$ `r round(TTE + HTE, 2)`

$\Delta CS = \alpha \left[ \ln u \right]_{P_w}^{P_1} =$ `r round(dCS,2)`

$\Delta PS = \frac{\beta}{\epsilon + 1} \left[ u^{\epsilon + 1} \right]_{P_w}^{P_1} =$ `r round(dPS,2)`

$\Delta Tariffs = (P_w^1 - P_w) M =$ `r round(dTariffs,2)`

$\Delta W =$ `r round(dCS + dPS + dTariffs,2)`

----


```{r echo=FALSE, warning=FALSE, message=FALSE}
plt5 <- ggplot(df, aes(x=x,y=y,color=curve)) + 
  geom_line(size=1.5) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  annotate("segment", x = 0, xend = q1, y = p1, yend = p1, linetype = 3) +
  annotate("segment", x = q1, xend = q1, y = p1, yend = 0, linetype = 3) +
  annotate("segment", x = q1_d, xend = q1_d, y = p1, yend = 0, linetype = 3) +
  annotate("segment", x = 0, xend = qw, y = pw, yend = pw, linetype = 2) +
  annotate("segment", x = qw_d, xend = qw_d, y = pw, yend = 0, linetype = 2) +
  annotate("segment", x = par_alpha, xend = par_alpha, y = 1, yend = 0, linetype = 2) +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0,2)) +
  xlab("quantity") +
  ylab("price") +
  labs(title="Demand and Supply",
       subtitle = "Impact of a 20% Tariff") +
  annotate("text", x=350, y=pw + .05, label="p[w]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=p1 + .05, label="p[1]", parse = TRUE, size = 5.2) +
  annotate("text", x = qw, y=0.1, label = "q[w]", parse = TRUE, size = 5.2) +
  annotate("text", x = q1, y=0.1, label = "q[1]", parse = TRUE, size = 5.2) +
  annotate("text", x = qw_d, y=0.1, label = "q[w][d]", parse = TRUE, size = 5.2) +
  annotate("text", x = q1_d, y=0.1, label = "q[1][d]", parse = TRUE, size = 5.2)
  
print(plt5)  
```



## Expected Outcome of Tariffs

Basic micro tells us..

```{r echo=FALSE, warning=FALSE, message=FALSE}


supply <- Hmisc::bezier(x = c(1, 8, 9),
                        y = c(2, 4.5, 9)) %>%
  as_data_frame()

demand <- Hmisc::bezier(x=c(1, 3, 9),
                        y=c(9, 3, 2)) %>%
  as_data_frame()

plot <- ggplot(mapping = aes(x,y)) + 
  geom_path(data = supply, color = "Blue", size = 1) + 
  geom_path(data = demand, color = "Orange", size = 1)+
  coord_equal()
plot
```
## what theory tells us..

# Equilibrium (no trade)

```{r echo=FALSE, warning=FALSE, message=FALSE}

func_supply <- approxfun(supply$x, supply$y, rule=2)
func_demand <- approxfun(demand$x, demand$y, rule=2)

eql_x <- uniroot(function(x) func_supply(x)-func_demand(x), c(0,10)) 
x_int <- eql_x$root
y_int <- func_demand(x_int)


plot + geom_hline(yintercept=y_int) + geom_vline(xintercept=x_int)
```

# Equilibrium (with trade)

```{r echo=FALSE, warning=FALSE, message=FALSE}
plot + geom_hline(yintercept = 2.5)
```

# A Tariff is Introduced
```{r echo=FALSE, warning=FALSE, message=FALSE}
plot + geom_hline(yintercept = 3.5)
```

## Net Loss of Tariff
``````{r echo=FALSE, warning=FALSE, message=FALSE}
plot + geom_hline(yintercept=2.5) + geom_hline(yintercept=3.5)

#we know the y value.. need x
```

## System Equations

Domestic Demand: $$Q_d=\alpha(P)^\eta$$

Domestic Supply: $$Q_s=\beta(P)^\epsilon$$

Import Demand (the excess of domestic demand over domestic supply): $$M_d=\alpha(P)^\eta-\beta(P)^\epsilon$$

Import Supply: $$M_s=Y(P_w)^\theta$$

Internal Price: $$P=(1+t)*P_w$$

# Assumptions
Some text about the linear vs. non linear general approach (the presentation is a simplifcation to make it easy to understand)

# Calibration
Starting values given 

# The Basic Model
In which changes can lead to incorrect outcomes because the simplification only makes simulation results true locally

## Intro GSIM
Given the limitations of the simple model, GSIM is an alternative based on a 4x4 region trade region that assumes a linear form of the system equations.

# Intro a Tariff
And show how simulation compares here

## Simulation results

# Welfare outcomes?

## Future Work
Could be extended to model 11x11 provinces (or 12x12 with ROW) to look at provincial trade flows. Some related work being done internally, and some outside work (Stats Can Hexagon Paper) where gravity models suggest that trade flows could potentially be quite a bit higher and are impeded by barriers. What would we find if we relaxed some trade barriers?

## Newton's Method
```{r echo=FALSE, warning=FALSE, message=FALSE}
a <- 0.25
b <- -1
c <- 3
d <- -1

x <- seq(-1, 3, 0.1)
y <- a * (x - b)^c + d

x00 <- b + (-d/a)^(1/c)

x0 <- 2
y0 <-  a * (x0 - b)^c + d

dydx <- c * a * (x0 - b)^(c-1)

yint <- y0 - dydx * x0

y1 <- 0
x1 <- x0 + (y1 - y0) / dydx

x2 <- 0.98 * x0
y2 <- 0.9 * y0 + dydx * (x2 - x0)

x3 <- 1.1 * x1
y3 <- 0.9 * y0 + dydx * (x3 - x0)


df <- data.frame(x = x, y = y)
plt <- ggplot(df, aes(x,y)) + 
  geom_line(size=1.2) + 
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_abline(intercept = yint, slope = dydx) + 
  annotate("segment", x = 0, xend = x0, y = y0, yend = y0, linetype = 2) +
  annotate("segment", x = x0, xend = x0, y = y0, yend = 0, linetype = 2) +
  geom_point(data=data.frame(x=c(x00, x1, x0, x0), y=c(0,0,0,y0)), size=1.5, color="red") +
  annotate("text", x = x0, y=0.5, label="x[0]", parse=TRUE) +
  annotate("text", x = x1, y=0.5, label="x[1]", parse=TRUE) +
  annotate("text", x = x00, y=0.5, label="x*", parse=FALSE) + 
  annotate("segment", x= x2, xend = x3, y = y2, yend = y3, arrow=arrow(length = unit(0.2, "cm")))
plt

```

