---
title: "Modelling Trade"
subtitle: "A Partial Equilibrium Approach"
author: "Shahzia Teja"
date: "July 9, 2019"
output: 
  revealjs::revealjs_presentation:
    theme: beige
    highlight: haddock
    css: style.css
    self_contained: false
    reveal_plugins: ["chalkboard"]
    reveal_options:
      chalkboard:
        theme: chalkboard
        toggleNotesButton: true
      slide_level: 2
      slideNumber: true
      previewLinks: true
      progress: true
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
options(stringsAsFactors = FALSE, scipen = 20)
library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)
```

# Ceteris Paribus {data-background-iframe="https://vi.unctad.org/tpa/web/vol1/vol1home.html"}

## Partial Equilibrium for Trade \& Regional Analysis

Partial Equilibrium models allow us to examine the effects of economic shocks in a parsimonious modelling environment with few data requirements or assumptions.

In this world, a specific market clears without regard to prices and quantity changes in other markets. In other words, all other prices, quantities and income are assumed to be fixed. Some of the advantages of this approach are obvious, including the ability to create and run models quickly, as well as producing intuitively appealing results.

## Tariffs as a Shock

When countries impose trade barriers, in this case in the form of tariffs, prices and quantities of implicated goods can be expected to respond.

We assume a large open economy with downward sloping import demand curves (this implies the country's demand affects prices). In general, an importing country is assumed to have prices above the world price, since the alternative would imply no trade takes place. Buyers are assumed to be indifferent between domestic and foreign-produced substitutes.

# Simplest Partial Equilibrium Model

## System Functions

Domestic Demand: $Q_d=\alpha P^\eta$

Domestic Supply: $Q_s=\beta P^\epsilon$

Import Demand (the excess of domestic demand over domestic supply): $M_d=\alpha P^\eta-\beta P^\epsilon$

Import Supply: $M_s=\gamma P_w^\theta$

Internal Price: $P=(1+t) P_w$

## Autarky

Without imports, domestic demand equals domestic supply.

$Q_d = Q_s$ or $\alpha P^\eta = \beta P^\epsilon$

which implies $\alpha / \beta = P^{\epsilon - \eta}$

This can be solved for:
$$P_0 = (\alpha / \beta) ^ \frac{1}{\epsilon - \eta}$$
$$Q_0 = \alpha P_0^\eta = \alpha (\alpha / \beta) ^ \frac{\eta}{\epsilon - \eta} = \alpha ^ \frac{\epsilon}{\epsilon - \eta} \beta ^ \frac{-\eta}{\epsilon - \eta}$$

## Free Trade

In this case, import demand equals import supply...

$M_d = M_s$ or $\alpha P^\eta - \beta P^\epsilon = \gamma P_w^\theta$

but $P = (1 + t) P_w$ which implies:

$\alpha (1 + t)^\eta P_w^\eta - \beta (1 + t)^\epsilon P_w^\epsilon = \gamma P_w^\theta$


## Autarky Illustrated

```{r echo=FALSE, warning=FALSE, message=FALSE}
# demand and supply
Qd <- function(p=1, alpha = 100, eta = -1) {
  alpha * p ^ eta
}

Pd <- function(q = 100, alpha = 100, eta = -1) {
  (q/alpha)^(1/eta)
}

Qs <- function(p = 1, beta = 100, epsilon = 3) {
  beta * p ^ epsilon
}


par_alpha <- 7480.06
par_beta  <- 1182.38
par_gamma <- par_alpha - par_beta

par_eta     <- -1
par_epsilon <- 3
par_theta   <- 10

# no trade eq'm
p0 <- (par_alpha / par_beta) ^ (1 / (par_epsilon - par_eta))
q0 <- par_alpha * p0 ^ par_eta

# benchmark equilibrium
pars <- c(a = par_gamma, b = par_theta - par_eta, c= par_beta, d = par_epsilon - par_eta, e = par_alpha)

f_bm <- function(pw, pars) {
  unname(pars['a'] * pw ^ pars['b'] + pars['c'] * pw ^ pars['d'] - pars['e'])
}

rslt <- uniroot(f_bm, c(0.1,2), pars, tol=5e-10)
pw <- rslt$root

qw <- par_alpha * pw ^ par_eta
qw_d <- par_beta * pw ^ par_epsilon
qw_i <- qw - qw_d
```

Suppose: 

$\alpha =$ `r par_alpha`, $\beta =$ `r par_beta`, $\gamma =$ `r par_gamma`,

$\eta =$ `r par_eta`, $\epsilon =$ `r par_epsilon` and $\theta =$ `r par_theta` 

Ignoring tariffs for now...

This implies $P_0 =$ `r round(p0,2)` and $Q_0 =$ `r round(q0,2)`

Whereas in the benchmark case, 

$P^* = P_w = 1$ and $Q^* = Q_{Domestic} + Q_{Imported}$

= `r par_beta` + `r par_gamma` = `r par_alpha`

---

```{r echo=FALSE, warning=FALSE, message=FALSE}
p <- seq(0.01, 2, .01)
qd <- Qd(p, par_alpha)  # demand
qs <- Qs(p, par_beta)  # domestic supply
md <- qd - qs         # import demand

mx <- max(c(qd, qs))
mn <- min(c(qd, qs))

df1 <- data.frame(x = qd, y = p, curve = rep("demand", length(p)))
df2 <- data.frame(x = qs, y = p, curve = rep("supply", length(p)))
df3 <- data.frame(x = md, y = p, curve = rep("imports", length(p)))

df <- rbind(df1, df2, df3[df3$x >= 0,])

plt1 <- ggplot(df, aes(x=x,y=y,color=curve)) + 
  geom_line(size=1.5) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  annotate("segment", x = 0, xend = q0, y = p0, yend = p0, linetype = 3) +
  annotate("segment", x = q0, xend = q0, y = p0, yend = 0, linetype = 3) +
  annotate("segment", x = 0, xend = par_alpha, y = 1, yend = 1, linetype = 2) +
  annotate("segment", x = par_beta, xend = par_beta, y = 1, yend = 0, linetype = 2) +
  annotate("segment", x = qw_i, xend = qw_i, y = 1, yend = 0, linetype = 2) +
  annotate("segment", x = par_alpha, xend = par_alpha, y = 1, yend = 0, linetype = 2) +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0,2)) +
  xlab("quantity") +
  ylab("price") +
  labs(title="Demand, Supply and Import Demand",
       subtitle = "Import Demand is residual demand") +
  annotate("text", x=q0, y=0.1, label="q[0]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=1.05, label="p[w]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=p0 + .05, label="p[0]", parse = TRUE, size = 5.2) +
  annotate("text", x = par_beta, y=0.1, label = "q[Dom]", parse = TRUE, size = 5.2) +
  annotate("text", x = qw_i, y=0.1, label = "q[Imp]", parse = TRUE, size = 5.2) +
  annotate("text", x = par_alpha, y=0.1, label = "q[Tot]", parse = TRUE, size = 5.2)
print(plt1)  

```

## Impact of Moving from Autarky to Free Trade

Using demand and supply curves defined above, price falls from $P_0$ = `r round(p0, 2)` to $P_w$ = `r round(pw,2)`

## Consumer Surplus

$\Delta CS = -\int_{P_0}^{P_w} Q_d(u) du$

but $p_w$ < $p_0$, so, 

$\Delta CS = -\int_{P_w}^{P_0} -Q_d(u) du = \alpha \int_{P_w}^{P_0} u^\eta du$

if $\eta \neq -1$, this equals $\alpha \left[\frac{u^{\eta + 1}}{\eta + 1}\right]_{P_w}^{P_0}$ 

if $\eta = -1$, this equals $\alpha \left[\ln u \right]_{P_w}^{P_0}$


In this case, $\Delta CS =$ `r round(par_alpha * (log(p0) - log(pw)),2)`

---

```{r echo=FALSE, warning=FALSE, message=FALSE}
df <- rbind(df1, df2)

df4 <- data.frame(x = qd, y = p, curve="demand")
df4 <- subset(df4, x <= qw & (y >= pw & y <= p0))
df4 <- rbind(df4, df4[1:2,])
row.names(df4) <- 1:nrow(df4)
df4[nrow(df4)-1, 1] <- 0
df4[nrow(df4), 1] <- 0
df4[nrow(df4)-1, 2] <- p0
df4[nrow(df4), 2] <- 1

plt2 <- ggplot(df, aes(x=x,y=y,color=curve)) + 
  geom_line(size=1.5) +
  geom_polygon(data = df4, aes(x=x,y=y), color="grey", alpha=0.2, show.legend = FALSE) +
  geom_hline(yintercept = c(0, pw, p0)) +
  geom_vline(xintercept = 0) +
  annotate("segment", x= 350, xend = 350, y = p0 - 0.01, yend = pw + 0.01, arrow=arrow(length = unit(0.2, "cm"))) +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0,2)) +
  annotate("text", x=350, y=p0 + .05, label="p[0]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=pw - .05, label="p[w]", parse = TRUE, size = 5.2) +
  xlab("quantity") +
  ylab("price") +
  labs(title = "Increase in Consumer Surplus",
       subtitle = "Opening to trade makes consumers better off")

print(plt2)
```

## Producer Surplus

$\Delta PS = \int_{P_0}^{P_w} Q_s(u) du$

but $p_w$ < $p_0$, so, 

$\Delta PS = -\int_{P_w}^{P_0} Q_s(u) du = -\beta \int_{P_w}^{P_0} u^\epsilon du$

Since $\epsilon > 0$, this equals $-\beta \left[\frac{u^{\epsilon + 1}}{\epsilon + 1} \right]_{P_w}^{P_0}$


In this case, $\Delta PS =$ `r round(-par_beta * (p0^(par_epsilon+1) - pw^(par_epsilon+1))/(par_epsilon+1),2)`

---

```{r echo=FALSE, warning=FALSE, message=FALSE}
# producer surplus
df5 <- data.frame(x = qs, y = p, curve="supply")
df5 <- subset(df5, x <= qw & (y >= pw & y <= p0))
df5 <- rbind(df5, df5[1:2,])
row.names(df5) <- 1:nrow(df5)
df5[nrow(df5)-1, 1] <- 0
df5[nrow(df5), 1] <- 0
df5[nrow(df5)-1, 2] <- p0
df5[nrow(df5), 2] <- 1

plt3 <- ggplot(df, aes(x=x,y=y,color=curve)) + 
  geom_line(size=1.5) +
  geom_polygon(data = df5, aes(x=x,y=y), color="grey", alpha=0.2, show.legend = FALSE) +
  geom_hline(yintercept = c(0, pw, p0)) +
  geom_vline(xintercept = 0) +
  annotate("segment", x= 350, xend = 350, y = p0 - 0.01, yend = pw + 0.01, arrow=arrow(length = unit(0.2, "cm"))) +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0,2)) +
  annotate("text", x=350, y=p0 + .05, label="p[0]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=pw - .05, label="p[w]", parse = TRUE, size = 5.2) +
  xlab("quantity") +
  ylab("price") +
  labs(title = "Reduction in Producer Surplus",
       subtitle = "Opening to trade makes producers worse off")

print(plt3)
```

## Net Change in Welfare

In this case we have $\Delta CS + \Delta PS > 0$

In fact, $\Delta CS + \Delta PS =$ `r round( par_alpha * (log(p0) - log(pw))- par_beta * (p0^(par_epsilon+1) - pw^(par_epsilon+1))/(par_epsilon+1), 2)`

---

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Net gain in welfare
x <- seq(pw, p0, .01)
y <- Qd(x, par_alpha)
x1 <- seq(p0 - 0.01, pw, -0.01)
y1 <- Qs(x1, par_beta)
df6 <- data.frame(x = c(y,y1), y = c(x,x1), curve="demand")

plt4 <- ggplot(df, aes(x=x,y=y,color=curve)) + 
  geom_line(size=1.5) +
  geom_polygon(data = df6, aes(x=x,y=y), color="grey", alpha=0.2, show.legend = FALSE) +
  geom_hline(yintercept = c(0, pw, p0)) +
  geom_vline(xintercept = 0) +
  annotate("segment", x= 350, xend = 350, y = p0 - 0.01, yend = pw + 0.01, arrow=arrow(length = unit(0.2, "cm"))) +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0,2)) +
  annotate("text", x=350, y=p0 + .05, label="p[0]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=pw - .05, label="p[w]", parse = TRUE, size = 5.2) +
  xlab("quantity") +
  ylab("price") +
  labs(title = "Net Increase in Welfare",
       subtitle = "Consumers more better off than producers are worse off")

print(plt4)
```

# Impact of Tariff in Simple Model

## Prices \& Quantities

```{r echo=FALSE, warning=FALSE, message=FALSE}
# tariff equilibrium
t <- 0.2
pars <- c(a = par_gamma, 
          b = par_theta - par_eta, 
          c = par_beta * (1 + t) ^ par_epsilon, 
          d = par_epsilon - par_eta, 
          e = par_alpha * (1 + t) ^ par_eta)

rslt <- uniroot(f_bm, c(0.1,2), pars, tol=5e-10)
pw1 <- rslt$root

p1 <- (1 + t) * pw1
q1 <- par_alpha * p1 ^ par_eta
q1_d <- par_beta * p1 ^ par_epsilon
q1_i <- q1 - q1_d

v_before <- c(pw, pw, qw_i, qw_d, qw, 0)
v_after  <- c(pw1, p1, q1_i, q1_d, q1, (p1 - pw1) * q1_i)
v_change <- v_after - v_before
v_pct    <- ifelse(v_before > 0, v_change / v_before * 100, NA)
v_decs   <- c(3, 3, 2, 2, 2, 2)

c1 <- mapply(function(x,y) format(round(x,y), nsmall = y), v_before, v_decs)
c2 <- mapply(function(x,y) format(round(x,y), nsmall = y), v_after, v_decs)
c3 <- mapply(function(x,y) format(round(x,y), nsmall = y), v_change, v_decs)
c4 <- format(round(v_pct,2),nsmall = 2)

df_tab <- data.frame(Variables = c("Pw", "P", "Md", "Qs", "Qd", "Tariffs"),
                     Before = c1, After =  c2, Change = c3, Percent = c4)
kable(df_tab)

TTE <- -0.5 * (qw_i + q1_i) * (pw1 - pw) # Terms of Trade Effect

HTE <- 0.5 * (p1 - pw1) * (q1_i - qw_i) # Harberger Triangle Effect

WE  <- TTE + HTE # Welfare Effect

dCS <- -par_alpha * (log(p1) - log(pw))
dPS <- par_beta * (p1^(par_epsilon + 1) - pw^(par_epsilon + 1)) / (par_epsilon + 1)
dTariffs <- (p1 - pw1) * q1_i

p1_s <- 0.8 + 0.2 * par_theta * par_gamma / ((par_epsilon - par_eta) * par_beta + (par_theta - par_eta) * par_gamma)
```

Welfare effect:

$dW = -M dP_w + \frac{1}{2}(P - P_w) dM =$ `r round(TTE + HTE, 2)`

Terms of trade effect: $-M dP_w =$ `r round(TTE,2)`

Harberger triangle effect: $\frac{1}{2}(P - P_w) dM =$ `r round(HTE,2)`


## How Accurate is the approximation?

Total differential above implies $dW =$ `r round(TTE + HTE, 2)`

Comparing areas suggests:

$\Delta CS = \alpha \left[ \ln u \right]_{P_w}^{P_1} =$ `r round(dCS,2)`

$\Delta PS = \frac{\beta}{\epsilon + 1} \left[ u^{\epsilon + 1} \right]_{P_w}^{P_1} =$ `r round(dPS,2)`

$\Delta Tariffs = (P_w^1 - P_w) M =$ `r round(dTariffs,2)`

$\Delta W =$ `r round(dCS + dPS + dTariffs,2)`

---

```{r echo=FALSE, warning=FALSE, message=FALSE}
plt5 <- ggplot(df, aes(x=x,y=y,color=curve)) + 
  geom_line(size=1.5) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  annotate("segment", x = 0, xend = q1, y = p1, yend = p1, linetype = 3) +
  annotate("segment", x = q1, xend = q1, y = p1, yend = 0, linetype = 3) +
  annotate("segment", x = q1_d, xend = q1_d, y = p1, yend = 0, linetype = 3) +
  annotate("segment", x = 0, xend = qw, y = pw, yend = pw, linetype = 2) +
  annotate("segment", x = qw_d, xend = qw_d, y = pw, yend = 0, linetype = 2) +
  annotate("segment", x = par_alpha, xend = par_alpha, y = 1, yend = 0, linetype = 2) +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0,2)) +
  xlab("quantity") +
  ylab("price") +
  labs(title="Demand and Supply",
       subtitle = "Impact of a 20% Tariff") +
  annotate("text", x=350, y=pw + .05, label="p[w]", parse = TRUE, size = 5.2) +
  annotate("text", x=350, y=p1 + .05, label="p[1]", parse = TRUE, size = 5.2) +
  annotate("text", x = qw, y=0.1, label = "q[w]", parse = TRUE, size = 5.2) +
  annotate("text", x = q1, y=0.1, label = "q[1]", parse = TRUE, size = 5.2) +
  annotate("text", x = qw_d, y=0.1, label = "q[w][d]", parse = TRUE, size = 5.2) +
  annotate("text", x = q1_d, y=0.1, label = "q[1][d]", parse = TRUE, size = 5.2)
  
print(plt5)  
```

# Assumptions

## Linear or Non-linear?

<small>
From above, equilibrium occurs at the root of
$f(P_w, t) = \alpha (1 + t)^\eta P_w^\eta - \beta (1 + t)^\epsilon P_w^\epsilon - \gamma P_w^\theta$

We can observe that
$\frac{\partial f(P_w,t)}{\partial P_w} = \eta \alpha (1 + t)^\eta P_w^{\eta-1} - \epsilon \beta (1 + t)^\epsilon P_w^{\epsilon-1} - \theta \gamma P_w^{\theta-1}$
and
$\frac{\partial f(P_w,t)}{\partial t} = \eta \alpha (1 + t)^{\eta-1} P_w^\eta-1 - \epsilon \beta (1 + t)^{\epsilon-1} P_w^\epsilon$

So, a linear approximation (a plane) is given by
$\hat{f}(P_w, t) = f(P_w^0,t^0) + \frac{\partial f(P_w^0,t^0)}{\partial P_w}(P_w - P_w^0) + \frac{\partial f(P_w^0,t^0)}{\partial t}(t - t^0)$

$\hat{f}(P_w,t)$ should be close to $f(P_w, t)$ near $(P_w^0,t^0)$

Is the solution of $\hat{f}(P_w,t) = 0$ close to $f(P_w, t) = 0$?
</small>

## Using the Linear Approximation

```{r echo=FALSE, warning=FALSE, message=FALSE}
f0 <- function(Pw = 1, t =0) {
  par_alpha * (1 + t) ^ par_eta * Pw ^ par_eta -
    par_beta * (1 + t) ^ par_epsilon * Pw ^ par_epsilon -
      par_gamma * Pw ^ par_theta
}

# linear approximation

dfdPw <- function(Pw = 1, t = 0) {
  par_eta * par_alpha * (1 + t) ^ par_eta * Pw ^ (par_eta - 1) - 
    par_epsilon * par_beta * (1 + t) ^ par_epsilon * Pw ^ (par_epsilon - 1) -
      par_theta * par_gamma * Pw ^ (par_theta - 1)
}

dfdt <- function(Pw = 1, t = 0) {
  par_eta * par_alpha * (1 + t) ^ (par_eta - 1) * Pw ^ par_eta -
    par_epsilon * par_beta * (1 + t) ^ (par_epsilon - 1) * Pw ^ par_epsilon
}

nsteps <- 1
pn <- 1
t <- 0
t1 <- 0.2
h <- (t1 - t) / nsteps
for(i in 1:nsteps) {
  pn1 <- pn - f0(pn, t) / dfdPw(pn, t) - dfdt(pn, t) / dfdPw(pn,t) * h
  t <- t + h
  pn <- pn1
}

pw1s <- pn

nsteps <- 4
pn <- 1
t <- 0
t1 <- 0.2
h <- (t1 - t) / nsteps
for(i in 1:nsteps) {
  pn1 <- pn - f0(pn, t) / dfdPw(pn, t) - dfdt(pn, t) / dfdPw(pn,t) * h
  t <- t + h
  pn <- pn1
}
pw14s <- pn
```


$\hat{f}(P_w,t) = 0$ implies that $P_w = P_w^0 - \frac{f(P_w^0,t^0)}{\frac{\partial f}{\partial P_w}(P_w^0,t^0)} - \frac{\frac{\partial f}{\partial t}(P_w^0,t^0)}{\frac{\partial f}{\partial t}(P_w^0,t^0)}(t - t^0)$

The solution to the non-linear equation gives $P_w^1 =$ `r format(round(pw1,4),nsmall=4)`.

The approximation yields 1-step value of $P_w^{1s} =$ `r format(round(pw1s,4),nsmall=4)` and a 4-step value of $P_w^{4s} =$ `r format(round(pw14s,4),nsmall=4)`. 

The 1-step estimate is off by `r format(round((pw1s - pw1) / pw1 * 100,2),nsmall=2)`% and the 4-step by `r format(round((pw14s - pw1) / pw1 * 100,2),nsmall=2)`% 


## What about Trumpian Tariffs?

```{r echo=FALSE, warning=FALSE, message=FALSE}
t <- 1
pars <- c(a = par_gamma, 
          b = par_theta - par_eta, 
          c = par_beta * (1 + t) ^ par_epsilon, 
          d = par_epsilon - par_eta, 
          e = par_alpha * (1 + t) ^ par_eta)

rslt <- uniroot(f_bm, c(0.1,2), pars, tol=5e-10)
ptrump <- rslt$root

nsteps <- 1
pn <- 1
t <- 0
t1 <- 1
h <- (t1 - t) / nsteps
for(i in 1:nsteps) {
  pn1 <- pn - f0(pn, t) / dfdPw(pn, t) - dfdt(pn, t) / dfdPw(pn,t) * h
  t <- t + h
  pn <- pn1
}

ptrump1s <- pn

nsteps <- 4
pn <- 1
t <- 0
t1 <- 1
h <- (t1 - t) / nsteps
for(i in 1:nsteps) {
  pn1 <- pn - f0(pn, t) / dfdPw(pn, t) - dfdt(pn, t) / dfdPw(pn,t) * h
  t <- t + h
  pn <- pn1
}
ptrump4s <- pn
```

Let's try a 100% tariff in the simple model...

The non-linear solution gives the new world price $P_{trump} =$ `r format(round(ptrump,4),nsmall=4)`.

The resulting domestic price is \$`r format(round(2*ptrump,2), nsmall=2)` which is just below the $`r format(round(p0,2), nsmall=2)` price with no trade.

The approximation yields 1-step value of $P_{trump}^{1s} =$ `r format(round(ptrump1s,4),nsmall=4)` and a 4-step value of $P_{trump}^{4s} =$ `r format(round(ptrump4s,4),nsmall=4)`. 

The 1-step is off by `r format(round((ptrump1s - ptrump) / ptrump * 100,2),nsmall=2)`% and the 4-step by `r format(round((ptrump4s - ptrump) / ptrump * 100,2),nsmall=2)`%

So, perhaps be cautious when using linear approximations... or, take small steps...

# Global Simulation Analysis of Industry-Level Trade Policy (GSIM) 

## Intro to GSIM 

<small>
GSIM is a widely used modelling strategy for the partial equilibrium analysis of global trade policy changes developed by Joseph Francois and H. Keith Hall.

GSIM is a static, deterministic, single commodity bilateral trade model driven by export supply and bilateral import demand equations. Imports and exports are assumed to be a function of the world price after taking into account relevant bilateral trade taxes or subsidies. Since tariffs are bilateral and differ from country to country, changes in tariffs lead to changes in relative prices that drive differential changes in imports from various sources. Elasticities of substitution (Armington elasticities) determine the extent to which changes in relative prices lead to switches in the source of imports.

The model solves numerically to find market clearing prices such that global imports equate to global exports.

Francois and Hall provide 4x4 and 25x25 Excel based calculators with linearized versions of the model. We have ported a generalized version of those models to the *R* software environment for statistical computing and graphics. 
</small>

## GSIM Limitations

Because the partial equilibrium approach ignores other products that may be substitutes in consumption or production, losses and gains are potentially overestimated, as transfers of resources to or from other sectors are ignored. However, the approach has the useful advantage of allowing for a relatively rapid and transparent analysis of a wide range of trade policy issues with a minimum of data and computational requirements.

GSIM is ideal for preliminary work especially when data availability is limited.

## Recent GSIM Example

In September 2018, the U.S. Department of Agriculture used GSIM to estimate the level of gross trade damage caused by retaliatory tariffs to U.S. agricultural exports by commodity.

In the model, Chinese import demand for product i coming from the U.S. falls as the result of a retaliatory tariff. The level imported comes from the following equation: 

$M_{i,(US,CN)}^\tau = \left[ \frac{ \alpha_{i,(US,CN)} }{ P_{i,(US)}^*(1+\tau_{i,(US,CN)})} \right]^\sigma P_{i,(CN)}^{\sigma - 1} E_{i,(CN)}$


# Working with GSIM

## Solution Methods

Choose export prices for each region so that global demand equals global supply.

$$
F(\mathbf P) =
F \left( 
\begin{matrix}
P_{US}\\
P_{JP}\\
P_{EU}\\
P_{RW}
\end{matrix}
\right) =
\left(
\begin{matrix}
F_{US}\\
F_{JP}\\
F_{EU}\\
F_{RW}
\end{matrix}
\right)
\left(
\begin{matrix}
P_{US}\\
P_{JP}\\
P_{EU}\\
P_{RW}
\end{matrix}
\right)
=
\left(
\begin{matrix}
0\\
0\\
0\\
0\\
\end{matrix}
\right)
$$

We can use a multi-dimensional version of Newton's Method to improve a price vector...

## Recall Newton's Method

```{r echo=FALSE, warning=FALSE, message=FALSE}
a <- 0.25
b <- -1
c <- 3
d <- -1

x <- seq(-1, 3, 0.1)
y <- a * (x - b)^c + d

x00 <- b + (-d/a)^(1/c)

x0 <- 2
y0 <-  a * (x0 - b)^c + d

dydx <- c * a * (x0 - b)^(c-1)

yint <- y0 - dydx * x0

y1 <- 0
x1 <- x0 + (y1 - y0) / dydx

x2 <- 0.98 * x0
y2 <- 0.9 * y0 + dydx * (x2 - x0)

x3 <- 1.1 * x1
y3 <- 0.9 * y0 + dydx * (x3 - x0)


df <- data.frame(x = x, y = y)
plt <- ggplot(df, aes(x,y)) + 
  geom_line(size=1.2) + 
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_abline(intercept = yint, slope = dydx) + 
  annotate("segment", x = 0, xend = x0, y = y0, yend = y0, linetype = 2) +
  annotate("segment", x = x0, xend = x0, y = y0, yend = 0, linetype = 2) +
  geom_point(data=data.frame(x=c(x00, x1, x0, x0), y=c(0,0,0,y0)), size=1.5, color="red") +
  annotate("text", x = x0, y=0.5, label="x[0]", parse=TRUE) +
  annotate("text", x = x1, y=0.5, label="x[1]", parse=TRUE) +
  annotate("text", x = x00, y=0.5, label="x*", parse=FALSE) + 
  annotate("segment", x= x2, xend = x3, y = y2, yend = y3, arrow=arrow(length = unit(0.2, "cm")))
plt

```


## The Jacobian

<small>
$$J( \mathbf{P}) =
\left. 
\left[ 
\begin{matrix}
\frac{\partial F_{US}}{\partial P_{US}}  && \frac{\partial F_{US}}{\partial P_{JP}} && \cdots && \frac{\partial F_{US}}{\partial P_{RW}}\\
\frac{\partial F_{JP}}{\partial P_{US}}  && \frac{\partial F_{JP}}{\partial P_{JP}} && \cdots && \frac{\partial F_{JP}}{\partial P_{RW}}\\
\frac{\partial F_{EU}}{\partial P_{US}}  && \frac{\partial F_{EU}}{\partial P_{JP}} && \cdots && \frac{\partial F_{EU}}{\partial P_{RW}}\\
\frac{\partial F_{RW}}{\partial P_{US}}  && \frac{\partial F_{RW}}{\partial P_{JP}} && \cdots && \frac{\partial F_{RW}}{\partial P_{RW}}\\
\end{matrix}
\right] \;
\right\rvert_{\mathbf{P}=P_0} 
$$</small>

with

<small>
$$J_{ij} = \partial F_i / \partial P_j \approx 
\begin{cases}
(F(\cdots,P_j + h,\cdots) - F(P))/h\text{, difference quotient, or}\\
F(\cdots,P_j + h,\cdots) - F(P)\text{, finite difference} \\
\end{cases}$$
</small>


## Follow the Yellow Brick (_Jacobian_) Road 

A first order approximation to F around $P_0$ is given by

$F(\mathbf{P}) = F(\mathbf{P_0}) + J(\mathbf{P_0}) \mathbf{dP}$

As in the simple model above, we can solve for the price change that makes $F(\mathbf{P}) = \mathbf{0}$ as follows

$\mathbf{dP} = -J^{-1}(\mathbf{P_0}) F(\mathbf{P_0})$

So, take a step along $\mathbf{dP}$ and evaluate again...

# Simulation Results

```{r echo=FALSE, warning=FALSE, message=FALSE}
# GSIM 4x4 model 

# cross price effects
cross_price_effects <- function(x = rep(0,4), eq2, dT) {
  n <- ncol(eq2)
  rslt <- matrix(NA_real_, nrow=n, ncol = n)
  for(i in 1:n) {
    tmp <- matrix(1, nrow=n, ncol=1) %*% eq2[,i]
    tmp[n*1:n + 1:n - n] <- 0
    rslt[,i] <- tmp %*% (dT[,i] + x)
  }
  rslt
}

# change in demand
demand_effects <- function(x = rep(0,4), phi, eq3, chi, dT) {
  n <- length(x)
  rslt <- rep(NA_real_, n)
  for(i in 1:n) {
    rslt[i] <- (phi[i,] * eqn3[i,]) %*% (dT[i,] + x[i])
  }
  rslt + rowSums(phi * chi)
}

# calc numerical Jacobian
Jacobian <- function(x, eqn2, eqn3, phi, dT, xDemand) {
  J <- matrix(NA, nrow=4, ncol=4)
  h <- 1/32768
  
  for(i in 1:4) {
    x1 <- x
    x1[i] <- x[i] + h
    
    chi1 <- cross_price_effects(x1, eqn2, dT)
    
    dSupply1 <- x1 * Ex
    
    dDemand1 <- demand_effects(x1, phi, eqn3, chi1, dT)
    
    xDemand1 <- dDemand1 - dSupply1
    J[i,] <- (xDemand1 - xDemand ) / h
  }
  J
}

calcEqm <- function(x0, phi, eqn2, eqn3, dT) {
  x <- x0
  repeat {
    chi <- cross_price_effects(x, eqn2, dT)
    
    dSupply <- x * Ex
    
    dDemand <- demand_effects(x, phi, eqn3, chi, dT)
    
    xDemand <- dDemand - dSupply
    
    Jacob <- Jacobian(x, eqn2, eqn3, phi, dT, xDemand)
    
    x1 <- x - solve(Jacob) %*% xDemand / 1024 # 2^10 
    
    chi1 <- cross_price_effects(x1, eqn2, dT)
    
    dSupply1 <- x1 * Ex
    
    dDemand1 <- demand_effects(x1, phi, eqn3, chi1, dT)
    
    xDemand1 <- dDemand1 - dSupply1
    
    SSE <- sum(xDemand * xDemand)
    SSE1 <- sum(xDemand1 * xDemand1)
    if (SSE1 < SSE) x <- x1
    else break;
  }
  x
}
```


## Benchmark Case

```{r echo=FALSE, warning=FALSE, message=FALSE}
# elasticities
Em <- c(US = -1.25, JP = -1.25, EU = -1.25, RW = -1.25) # import demand

Ex <- c(US = 1.5, JP = 1.5, EU = 1.5, RW = 1.5) # export supply

Es <- c(US = 5, JP = 5, EU = 5, RW = 5) # substitution

# trade flows
M <- matrix(c(  0,  50, 200, 300,
              500,   0, 150, 200,
              300, 100, 200, 200,
               50, 100, 110,  20),
            nrow = 4, ncol=4, byrow = TRUE)

row.names(M) <- c("US", "JP", "EU", "RW")
colnames(M)  <- c("US", "JP", "EU", "RW")

# initial import tariffs 
T0 <- matrix(c(  1.00, 1.21, 1.41, 1.22,
                 1.37, 1.00, 1.31, 1.23,
                 1.32, 1.36, 1.00, 1.18,
                 1.57, 1.41, 1.25, 1.15),
              nrow = 4, ncol=4, byrow = TRUE)

row.names(T0) <- c("US", "JP", "EU", "RW")
colnames(T0)  <- c("US", "JP", "EU", "RW")

kable(M, caption = "Trade Flows")
```

<small>and</small>

```{r echo=FALSE,warning=FALSE, message=FALSE}
kable(T0, caption = "Tariffs")
```

# US-EU Bilateral Agreement

## Revised Tariffs US-EU

```{r warning=FALSE, message=FALSE}
T1 <- T0

T1["EU","US"] <- T1["US","EU"] <- 1 # US-EU trade agreement
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
kable(T1, caption = "Revised Tariffs")
```

## Equilibrium Prices US-EU

```{r echo=FALSE, warning=FALSE, message=FALSE}
# a vector of ones
i_agg <- matrix(1, nrow=nrow(M), ncol=1)

# Total imports and exports
Tot_imp <- t(i_agg) %*% M
row.names(Tot_imp) <- "Imports"

Tot_exp <- M %*% i_agg
colnames(Tot_exp) <- "Exports"

# import shares at internal prices
mult <- diag( c(1 / (t(i_agg) %*% (M * T0))))
row.names(mult) <- row.names(M)
colnames(mult) <- colnames(M)

theta <- (M * T0) %*% mult

# export shares at world prices
mult <- diag(1/c(M %*% i_agg))
row.names(mult) <- row.names(M)
colnames(mult) <- colnames(M)

phi <- mult %*% M

# own price elasticities
eqn3 <- theta %*% diag( Em ) - (1 - theta) %*% diag( Es )
colnames(eqn3) <- colnames(M)

# cross price elasticities
eqn2 <- theta %*% diag( Em + Es )
colnames(eqn3) <- colnames(M)

# reported prices
# x <- c(0.0792, -0.0316, 0.0480, -0.0184)
#
# try from zero...
x <- calcEqm(rep(0, 4), phi, eqn2, eqn3, T1/T0 - 1)

chi <- cross_price_effects(x, eqn2, T1/T0 - 1)
```

P = (`r paste(format(round(x,4),nsmall=2), sep=",")`)^T^

## Change in Trade Flows US-EU

```{r echo=FALSE, warning=FALSE, message=FALSE}
dPCT <- (x %*% t(i_agg) + T1/T0 - 1) * eqn3 + chi

M1 <- M * (1 + dPCT) * (1 + x %*% t(i_agg))

dM <- M1 - M

kable(dM, caption = "Impact of US-EU Tariff Elimination")
```


## Welfare Changes US-EU

```{r echo=FALSE, warning=FALSE, message=FALSE}
dP <- (1 + x %*% t(i_agg)) * T1/T0 -1

dCompP <- matrix(NA, nrow=1, ncol = ncol(dP))
row.names(dCompP) <- "Composite Price"
colnames(dCompP) <- c("US", "JP", "EU", "RW")
for(i in 1:ncol(dCompP)) {
  dCompP[1,i] <- sum(c(dP[,i] * theta[,i]))
}

dTariff <- M1 * (T1 - 1) - M * (T0 - 1)

dRev <- colSums(dTariff)

dCS <- c((0.5 * dCompP ^ 2) * colSums(M * T0) * sign(dCompP) * Em - colSums(M * T0) * dCompP)
names(dCS) <- c("US", "JP", "EU", "RW")

dPS <- c(x * Tot_exp *(1 + 0.5 * Ex * x))
names(dPS) <- c("US", "JP", "EU", "RW")

dW <- dPS + dCS + dRev

# make report

df_report <- data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"),
                        ProducerSurplus = format(round(dPS,1),nsmall=1),
                        ConsumerSurplus = format(round(dCS,1),nsmall=1),
                        TariffRevenue = format(round(dRev,1),nsmall=1),
                        NetWelfare = format(round(dW,1),nsmall=1))

kable(df_report, caption = "Impact of Tariff Changes", row.names = FALSE)
```


## Welfare Effects US-EU

```{r echo=FALSE, warning=FALSE, message=FALSE}
df_plot <- data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"), 
                      Component = "Producer Surplus", Value = dPS)

df_plot <- rbind(df_plot, 
                 data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"), 
                            Component = "Consumer Surplus", Value = dCS))

df_plot <- rbind(df_plot, 
                 data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"), 
                            Component = "Tariff Revenue", Value = dRev))

df_netW <- data.frame(Region=c("United States", "Japan", "European Union", "Rest of World"),
                      Component=NA,
                      Value=dW)

g <- ggplot(df_plot, aes(x=Region, y=Value, fill = Component)) + 
     geom_bar(stat = "identity") +
     scale_x_discrete(limits=c("United States", "Japan", "European Union", "Rest of World")) +
     geom_hline(yintercept = 0) + 
     geom_point(data=df_netW, size=2, show.legend = FALSE) +
     labs(x="", y="Welfare", title="Total Welfare Effects", subtitle = "By Component")

print(g)
```

# US-JP Bilateral Agreement

## Revised Tariffs US-JP

```{r warning=FALSE, message=FALSE}
T1 <- T0

T1["JP","US"] <- T1["US","JP"] <- 1 # US-EU trade agreement
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
kable(T1, caption = "Revised Tariffs US-JP")
```

## Equilibrium Prices US-JP

```{r echo=FALSE, warning=FALSE, message=FALSE}
# a vector of ones
i_agg <- matrix(1, nrow=nrow(M), ncol=1)

# Total imports and exports
Tot_imp <- t(i_agg) %*% M
row.names(Tot_imp) <- "Imports"

Tot_exp <- M %*% i_agg
colnames(Tot_exp) <- "Exports"

# import shares at internal prices
mult <- diag( c(1 / (t(i_agg) %*% (M * T0))))
row.names(mult) <- row.names(M)
colnames(mult) <- colnames(M)

theta <- (M * T0) %*% mult

# export shares at world prices
mult <- diag(1/c(M %*% i_agg))
row.names(mult) <- row.names(M)
colnames(mult) <- colnames(M)

phi <- mult %*% M

# own price elasticities
eqn3 <- theta %*% diag( Em ) - (1 - theta) %*% diag( Es )
colnames(eqn3) <- colnames(M)

# cross price elasticities
eqn2 <- theta %*% diag( Em + Es )
colnames(eqn3) <- colnames(M)

# reported prices
# x <- c(0.0792, -0.0316, 0.0480, -0.0184)
#
# try from zero...
x <- calcEqm(rep(0, 4), phi, eqn2, eqn3, T1/T0 - 1)

chi <- cross_price_effects(x, eqn2, T1/T0 - 1)
```

P = (`r paste(format(round(x,4),nsmall=2), sep=",")`)^T^

## Change in Trade Flows US-JP

```{r echo=FALSE, warning=FALSE, message=FALSE}
dPCT <- (x %*% t(i_agg) + T1/T0 - 1) * eqn3 + chi

M1 <- M * (1 + dPCT) * (1 + x %*% t(i_agg))

dM <- M1 - M

kable(dM, caption = "Impact of US-JP Tariff Elimination")
```


## Welfare Changes US-JP

```{r echo=FALSE, warning=FALSE, message=FALSE}
dP <- (1 + x %*% t(i_agg)) * T1/T0 -1

dCompP <- matrix(NA, nrow=1, ncol = ncol(dP))
row.names(dCompP) <- "Composite Price"
colnames(dCompP) <- c("US", "JP", "EU", "RW")
for(i in 1:ncol(dCompP)) {
  dCompP[1,i] <- sum(c(dP[,i] * theta[,i]))
}

dTariff <- M1 * (T1 - 1) - M * (T0 - 1)

dRev <- colSums(dTariff)

dCS <- c((0.5 * dCompP ^ 2) * colSums(M * T0) * sign(dCompP) * Em - colSums(M * T0) * dCompP)
names(dCS) <- c("US", "JP", "EU", "RW")

dPS <- c(x * Tot_exp *(1 + 0.5 * Ex * x))
names(dPS) <- c("US", "JP", "EU", "RW")

dW <- dPS + dCS + dRev

# make report

df_report <- data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"),
                        ProducerSurplus = format(round(dPS,1),nsmall=1),
                        ConsumerSurplus = format(round(dCS,1),nsmall=1),
                        TariffRevenue = format(round(dRev,1),nsmall=1),
                        NetWelfare = format(round(dW,1),nsmall=1))

kable(df_report, caption = "Impact of US-JP Tariff Changes", row.names = FALSE)
```


## Welfare Effects US-JP

```{r echo=FALSE, warning=FALSE, message=FALSE}
df_plot <- data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"), 
                      Component = "Producer Surplus", Value = dPS)

df_plot <- rbind(df_plot, 
                 data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"), 
                            Component = "Consumer Surplus", Value = dCS))

df_plot <- rbind(df_plot, 
                 data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"), 
                            Component = "Tariff Revenue", Value = dRev))

df_netW <- data.frame(Region=c("United States", "Japan", "European Union", "Rest of World"),
                      Component=NA,
                      Value=dW)

g <- ggplot(df_plot, aes(x=Region, y=Value, fill = Component)) + 
     geom_bar(stat = "identity") +
     scale_x_discrete(limits=c("United States", "Japan", "European Union", "Rest of World")) +
     geom_hline(yintercept = 0) + 
     geom_point(data=df_netW, size=2, show.legend = FALSE) +
     labs(x="", y="Welfare", title="Total Welfare Effects US-JP", subtitle = "By Component")

print(g)
```


# US Unilateral Action

## Revised Tariffs US Alone

```{r warning=FALSE, message=FALSE}
T1 <- T0

T1["JP","US"] <- T1["EU","US"] <- T1["RW","US"] <- 1
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
kable(T1, caption = "Revised Tariffs US Alone")
```

## Equilibrium Prices US Alone

```{r echo=FALSE, warning=FALSE, message=FALSE}
# a vector of ones
i_agg <- matrix(1, nrow=nrow(M), ncol=1)

# Total imports and exports
Tot_imp <- t(i_agg) %*% M
row.names(Tot_imp) <- "Imports"

Tot_exp <- M %*% i_agg
colnames(Tot_exp) <- "Exports"

# import shares at internal prices
mult <- diag( c(1 / (t(i_agg) %*% (M * T0))))
row.names(mult) <- row.names(M)
colnames(mult) <- colnames(M)

theta <- (M * T0) %*% mult

# export shares at world prices
mult <- diag(1/c(M %*% i_agg))
row.names(mult) <- row.names(M)
colnames(mult) <- colnames(M)

phi <- mult %*% M

# own price elasticities
eqn3 <- theta %*% diag( Em ) - (1 - theta) %*% diag( Es )
colnames(eqn3) <- colnames(M)

# cross price elasticities
eqn2 <- theta %*% diag( Em + Es )
colnames(eqn3) <- colnames(M)

# reported prices
# x <- c(0.0792, -0.0316, 0.0480, -0.0184)
#
# try from zero...
x <- calcEqm(rep(0, 4), phi, eqn2, eqn3, T1/T0 - 1)

chi <- cross_price_effects(x, eqn2, T1/T0 - 1)
```

P = (`r paste(format(round(x,4),nsmall=2), sep=",")`)^T^

## Change in Trade Flows US Alone

```{r echo=FALSE, warning=FALSE, message=FALSE}
dPCT <- (x %*% t(i_agg) + T1/T0 - 1) * eqn3 + chi

M1 <- M * (1 + dPCT) * (1 + x %*% t(i_agg))

dM <- M1 - M

kable(dM, caption = "Impact of US Tariff Elimination")
```


## Welfare Changes US Alone

```{r echo=FALSE, warning=FALSE, message=FALSE}
dP <- (1 + x %*% t(i_agg)) * T1/T0 -1

dCompP <- matrix(NA, nrow=1, ncol = ncol(dP))
row.names(dCompP) <- "Composite Price"
colnames(dCompP) <- c("US", "JP", "EU", "RW")
for(i in 1:ncol(dCompP)) {
  dCompP[1,i] <- sum(c(dP[,i] * theta[,i]))
}

dTariff <- M1 * (T1 - 1) - M * (T0 - 1)

dRev <- colSums(dTariff)

dCS <- c((0.5 * dCompP ^ 2) * colSums(M * T0) * sign(dCompP) * Em - colSums(M * T0) * dCompP)
names(dCS) <- c("US", "JP", "EU", "RW")

dPS <- c(x * Tot_exp *(1 + 0.5 * Ex * x))
names(dPS) <- c("US", "JP", "EU", "RW")

dW <- dPS + dCS + dRev

# make report

df_report <- data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"),
                        ProducerSurplus = format(round(dPS,1),nsmall=1),
                        ConsumerSurplus = format(round(dCS,1),nsmall=1),
                        TariffRevenue = format(round(dRev,1),nsmall=1),
                        NetWelfare = format(round(dW,1),nsmall=1))

kable(df_report, caption = "Impact of US Tariff Changes", row.names = FALSE)
```


## Welfare Effects US Only

```{r echo=FALSE, warning=FALSE, message=FALSE}
df_plot <- data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"), 
                      Component = "Producer Surplus", Value = dPS)

df_plot <- rbind(df_plot, 
                 data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"), 
                            Component = "Consumer Surplus", Value = dCS))

df_plot <- rbind(df_plot, 
                 data.frame(Region = c("United States", "Japan", "European Union", "Rest of World"), 
                            Component = "Tariff Revenue", Value = dRev))

df_netW <- data.frame(Region=c("United States", "Japan", "European Union", "Rest of World"),
                      Component=NA,
                      Value=dW)

g <- ggplot(df_plot, aes(x=Region, y=Value, fill = Component)) + 
     geom_bar(stat = "identity") +
     scale_x_discrete(limits=c("United States", "Japan", "European Union", "Rest of World")) +
     geom_hline(yintercept = 0) + 
     geom_point(data=df_netW, size=2, show.legend = FALSE) +
     labs(x="", y="Welfare", title="Total Welfare Effects US Alone", subtitle = "By Component")

print(g)
```

# Future Work

## Provincial Barriers to Trade

Could easily be extended to model 11x11 provinces (including 10 provinces and Rest of World) to look at provincial trade flows. 

Some related work being done internally, and some outside work (Stats Can Hexagon Paper comes to mind) where gravity models suggest that trade flows could potentially be quite a bit higher and are impeded by barriers. 

What would we find if we relaxed some trade barriers?

# The End
